A fast and Low-Overhead Technique to Secure Programs Against Integer Overflows

### 2012-11-06 doc/Manuscript/CGO/paper.tex
Added comments from the reviewers to the paper.

\title{A fast and Low-Overhead Technique to Secure Programs Against
Integer Overflows}
%\title{Combining Static and Dynamic Analyses to Secure Programs against
%Integer Overflows}
%\subtitle{A case study of the Dinamica EGO toolkit}

\authorinfo{Removed due to blind review}
           {No affiliation given}
           {No e-mail given}

@@

% Why there is still the need to devise efficient methods to eliminate them.
Programming languages such as Ada or Lisp can be customized to throw exceptions
whenever integer overflows are detected.
Furthermore, there exist recent work proposing to instrument binaries derived
from C, C++ and Java programs to detect the occurrence of overflows
dynamically~\cite{Brumley07,Dietz12}.
Thus, if an overflow happens, then the instrumented program can take some
action, such as to log the event, or to terminate the program.
However, this safety comes with a price: arithmetic operations need to be
surveilled, and the runtime checks cost time.
Zhang {\em et al.}~\cite{Zhang09} have proposed to eliminate some of this
overhead via a tainted flow analysis.
We have a similar goal, yet, our approach is substantially different.

% Our contribution: range analysis.
This paper describes the range analysis algorithm that
we have developed to eliminate overflow checks in instrumented programs.
As we show in Section~\ref{sec:range}, our algorithm has three core insights.
Firstly, in Section~\ref{sub:micro} we propose a three-phases approach to extract
information from comparisons between variables, e.g., $x < y$.
Previous algorithms either deal with these comparisons via
expensive relational analyses~\cite{Cousot78,Lakhdar11,Mine06}, or only
consider comparisons between variables and
constants~\cite{Mahlke01,Patterson95,Stephenson00}.
Secondly, in Section~\ref{sub:prop} we show how we rely on strongly connected
components to achieve speed and precision.
It is well-known that this technique is effective in speeding up constraint
resolution~\cite[Sec 6.3]{Nielson99}.
However, given our three-phases approach, we improve precision
by solving strong components in topological order.
Finally, in Section~\ref{sub:splitting} we propose a new program representation that
is more precise than other intermediate forms used to solve range
analysis sparsely.
This new live range splitting strategy is only valid if the instrumented
program terminates whenever an integer overflow is detected.
If we cannot rely on this guarantee, then our more conservative live range splitting
strategy produces the program representation that Bodik
{\em et al.}~\cite{Bodik00} call Extended Static Single Assignment
form.
In Section~\ref{sub:range} we show that an interprocedural implementation of our
algorithm analyzes programs with half-a-million lines of code in less than ten
seconds.
Furthermore, the speed and memory consumption of our range analysis grows linearly
with the program size.

@@

% An example of vulnerability.
Figure~\ref{fig:ex_buffer_overflow} illustrates an example of integer
overflow vulnerability.
This vulnerability allows an adversary to perform a buffer overrun attack
on a program that is apparently guarded against this type of exploit.
For simplicity, we will use the C datatype \texttt{char}, which represents
8-bit long numbers.
Examples of actual vulnerabilities, with larger primitive types, are given
by Brumley {\em et al.}~\cite{Brumley07}, Dietz {\em et al.}~\cite{Dietz12}, and
Zhang {\em et al.}~\cite{Zhang09}.
The function \texttt{read\_matrix} reads a matrix of bytes from a file.
This matrix is stored as a linear sequence of bytes in a vector \texttt{out}.
The first and second bytes in the file, e.g., \texttt{w} and \texttt{h},
determine the number of rows and columns in the matrix.
If the \texttt{char} datatype could represent arbitrarily long numbers, than this
function would be safe against buffer overruns, because it would only write data
on allocated memory.
However, if, for instance, we have that \texttt{w}$= 16$ and \texttt{h}$= 16$,
then \texttt{s}$= 0$, as pointed before.
In this case, a buffer of length zero would be allocated, and all the data
stored in the file would overwrite memory in the stack of activation records.
By carefully crafting an input file, an adversary can, in this way, overwrite
the return address of \texttt{read\_matrix}, diverting program's execution to
one of \texttt{libc}'s function, such as \texttt{telnet}, for instance.

@@

From the lattice $\varset{Z}$ we define the product lattice
$\varset{Z}^2$ as follows:
%
\begin{equation*}
\varset{Z}^2 = \{ \emptyset \} \cup \{[z_1, z_2] | \ z_1,z_2 \in \varset{Z},
\ z_1 \leq z_2, \  -\infty < z_2 \}
\end{equation*}
%
This interval lattice is partially ordered by the subset relation, which we
denote by ``$\sqsubseteq$".
Range intersection, ``$\sqcap$", is defined by:
\[
[a_1, a_2] \sqcap [b_1, b_2] =
\begin{cases}
[\mbox{max}(a_1, b_1), \mbox{min}(a_2, b_2)], \ \mbox{if} \ a_1 \leq b_1 \leq a_2  \\ \ \ \ \ \mbox{or} \ b_1 \leq a_1 \leq b_2 \\
[a_1, a_2] \sqcap [b_1, b_2] = \emptyset, \ \mbox{otherwise}
\end{cases}
\]
And range union, ``$\sqcup$", is given by:
\[
[a_1, a_2] \sqcup [b_1, b_2] = [\mbox{min}(a_1, b_1), \mbox{max}(a_2, b_2)]
\]

Given an interval $\iota = [l, u]$, we let $\lb{\iota} = l$, and
$\ub{\iota} = u$, where $\lb{\iota}$ is the lower bound and $\ub{\iota}$ is
the upper bound of a variable.
We let \varset{V} be a set of constraint variables, and
$I: \varset{V} \mapsto \varset{Z}^2$ a
mapping from these variables to intervals in $\varset{Z}^2$.
Our objective is to solve a constraint system $C$, formed by constraints such
as those seen in Figure~\ref{fig:eval_function}(left).
We let the $\phi$-functions be as defined by Cytron
{\em et al.}~\cite{Cytron91}: they join different variable names into a single
definition.
Figure~\ref{fig:eval_function}(right) defines a valuation function $e$ on the
interval domain.
Armed with these concepts, we define the range analysis problem as follows:

\begin{definition}
\label{def:rcp}
\textsc{Range Analysis Problem} \\
\textbf{Input:} a set $C$ of constraints ranging over a set \varset{V} of
variables. \\
\textbf{Output:} a mapping I such that, for any variable
$V \in \varset{V}$, e(V) = I[V].
\end{definition}

@@

% The macro algorithm
Our range analysis algorithm works in a number of steps.
Firstly, we convert the program to an intermediate representation that gives
us subsidies to perform a sparse analysis.
We have tested our algorithm with two different representations, as we discuss
in Section~\ref{sub:splitting}.
Secondly, we extract constraints from the program representation.
Thirdly, we build a constraint graph, following the strategy pointed by
Su and Wagner~\cite{Su05}.
However, contrary to them, in a next phase we find the strongly connected
components in this graph, collapse them into super-nodes, and sort the
resulting digraph topologically.
Finally, for each strong component, we apply a three-phases approach to determine
the ranges of the variables.

\noindent
\textbf{Building the Constraint Graph. }
Given a set $\varset{C}$ of constraints, which define and/or use constraint
variables from a set $\varset{V}$, we build a constraint graph
$G = (\varset{C} \cup \varset{V}, E)$.
The vertices in \varset{C} are the {\em constraint nodes}, and the
nodes in \varset{V} are the {\em variable nodes}.
If $V \in \varset{V}$ is used in constraint $C \in \varset{C}$, then
we create an edge $\overrightarrow{VC}$.
If constraint $C \in \varset{C}$ defines variable $\varset{V} \in V$, then we
create an edge $\overrightarrow{CV}$.
Figure~\ref{fig:ex_graph} shows the constraint graph that we build for the
program in Figure~\ref{fig:ex_eSSA_cgo}(b).
Our algorithm introduces the notion of {\em future ranges}, which we use to extract
range information from comparisons between variables.
In Section~\ref{sub:splitting} we explain how futures are created.
If $V$ is used by $C$ as the input of a future range, then the edge from
$V$ to $C$ represents what Ferrante {\em et al.} call a {\em control
dependence}~\cite[p.323]{Ferrante87}.
We use dashed lines to represent these edges.
All the other edges denote {\em data dependences}~\cite[p.322]{Ferrante87}.
As we will show later, control dependence edges increase the precision of our
algorithm to solve future bounds.

@@

\subsection{A Three-Phases Approach to Solve Strong Components}
\label{sub:micro}

We determine the ranges for each strongly connected component in three phases.
First we determine the growth pattern of each variable in the component via
widening.
In the second step, once we know in which directions the values stored in each
variable might grow, we replace futures by actual bounds.
Finally, a narrowing phase starting from conditional tests improves the
precision of our results.

\noindent
\textbf{Widening: } we start solving constraints by determining how each program
variable might grow.
For instance, if a variable is only updated by sums with positive numbers, then
it only grows up.
If, instead, a variable is only updated by sums with negative numbers, then it
grows down.
Some variables can also grow in both directions.
We discover these growth patterns by abstractly interpreting the constraints that
constitute the strongly connected component.
We ensure termination via a widening operator.
Our implementation uses {\em jump-set widening}, which is typically used in
range analysis~\cite[p.228]{Nielson99}.
This operator is a generalization of Cousot and Cousot's original widening
operator~\cite{Cousot77}, which we describe below:
%
\begin{equation*}
I[Y] =
\begin{cases}
\mbox{if} \ I[Y] = [\bot, \bot] \ \mbox{then} \ e(Y) \\
\mbox{elif} \ \lb{e(Y)} < \lb{I[Y]}  \ \mbox{and} \ \ub{e(Y)} > \ub{I[Y]} \ \mbox{then} \ [-\infty, \infty] \\
\mbox{elif} \ \lb{e(Y)} < \lb{I[Y]} \ \mbox{then} \ [-\infty, \ub{I[Y]}] \\
\mbox{elif} \ \ub{e(Y)} > \ub{I[Y]} \ \mbox{then} \ [\lb{I[Y]}, \infty]
\end{cases}
\end{equation*}
%
We let $\lb{[l, u]} = l$ and $\ub{[l, u]} = u$.
The map $I$ and the abstract evaluation function $e$ are determined as in
Figure~\ref{fig:eval_function}.
We have a different implementation of $e$ for each operation that the
target programming language provides.
Our current LLVM implementation has 18 different instances of $e$, including
signed and unsigned addition, subtraction, multiplication and division, plus
truncation, the bitwise integer operators and $\phi$-functions.

@@

We have classified the bounds estimated by the static analysis into four
categories.
The first category, called $1$, contains tight bounds:
during program execution, the variable has been assigned an upper,
or lower limit, that equals the limit inferred statically.
The second category, called $n$, contains the bounds that are
within twice the value inferred statically.
For instance, if the range analysis estimates that a variable $v$ is in the
range $[0, 100]$, and during the execution the dynamic profiler finds that
its maximum value is $51$, then $v$ falls into this category.
The third category, $n^2$, contains variables whose actual value is within
a quadratic factor of the estimated value.
In our example, $v$'s upper bound would have to be at most $10$ for it to
be in this category.
Finally, the fourth category contains variables whose estimated value lays
outside a quadratic factor of the actual value.
We call this category {\em imprecise}, and it contains mostly the limits that
our static analysis has marked as either $+\infty$ or $-\infty$.
As we see in Figure~\ref{fig:precision}, 54.11\% of the lower limits that
we have estimated statically are exact.
Similarly, 51.99\% of our upper bounds are also tight.
The figure also shows that, on average, 37.39\% of our lower limits are
imprecise, and 35.40\% of our upper limits are imprecise.
This result is on pair with those obtained by more costly analysis, such as
Stephenson {\em et al.}'s~\cite{Stephenson00}.

@@

We have executed the instrumented programs of the integer benchmarks of SPEC
2006 CPU to probe the overhead imposed by our instrumentation.
These programs have been executed with their ``test" input sets.
We have not been able to run the binary that LLVM produces for SPEC's
\texttt{gcc} in our environment, even without any of our passes, due to an
incompatible \texttt{ctype.h} header.
In addition, we have not been able to collect the statistics regarding
the overflows occurred in SPEC's \texttt{bzip2}, because the log file
has became too large, due to the occurrence of more than 3.000.000.000 
overflows. Thus, the machine ran out of memory.
Figure~\ref{fig:unprunedTable} shows the percentage of instructions that
we instrument, without the intervention of the range analysis.
The number of instrumented instructions is relatively low, compared to the
total number of instructions, because we only instrument six different
LLVM bitcodes, in a set of 57 opcodes, not counting intrinsics.
Figure~\ref{fig:unprunedTable} also shows how many instructions have caused
overflows.
On the average, 4.90\% of the instrumented sites have caused integer overflows.

@@

%Prunning analysis
\begin{figure}[t!]
\begin{center}
\begin{small}
\begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}}|l|r|r|r|r|r|}
\hline
Benchmark & \#II & \#E & \#I/\#E & \#U &
\#I/\#U \\ \hline
470.lbm & 1.142 & 4 & 99,65\% & 4 & 99,65\% \\ \hline
433.milc & 1.573 & 1.070 & 31,98\% & 1.065 & 32,29\% \\ \hline
444.namd & 2.965 & 2.900 & 2,19\% & 2.900 & 2,19\% \\ \hline
447.dealII & 37.234 & 29.870 & 19,78\% & 28.779 & 22,71\% \\ \hline
450.soplex & 3.268 & 2.927 & 10,43\% & 2.897 & 11,35\% \\ \hline
464.h264ref & 13.548 & 11.342 & 16,59\% & 11.301 & 16,28\% \\ \hline
473.astar & 903 & 808 & 10,52\% & 806 & 10,74\% \\ \hline
458.sjeng & 2.483 & 2.354 & 5,20\% & 2.190 & 11,80\% \\ \hline
429.mcf & 165 & 164 & 0,61\% & 164 & 0,61\% \\ \hline
471.omnetpp & 1.533 & 1.313 & 14,35\% & 1.313 & 14,35\% \\ \hline
403.gcc & 16.330 & 15.282 & 6,42\% & 15.110 & 7,47\% \\ \hline
445.gobmk & 13.947 & 12.563 & 9,92\% & 12.478 & 10,53\% \\ \hline
462.libquantum & 943 & 820 & 13,04\% & 817 & 13,36\% \\ \hline
401.bzip2 & 2.350 & 1.966 & 16,34\% & 1.966 & 16,34\% \\ \hline
456.hmmer & 3.963 & 3.346 & 15,57\% & 3.304 & 16,63\% \\ \hline
Total (Average) & 6.823 & 5.779 & 18,17\% & 5.676 & 19,09\% \\ \hline
\end{tabular*}
\end{small}
\end{center}
\caption{\label{fig:prunningTable}
Instrumentation library with support of static range analysis.
\#II: number of instructions that have been instrumented without range
analysis.
\#E: number of instructions instrumented in the e-SSA form program.
\#U: number of instructions instrumented in the u-SSA form program.}
\end{figure}

Figure~\ref{fig:classificationTable} shows how our range analysis
classifies instructions.
Out of all the 102,790 instructions that we have instrumented in SPEC, 3.92\% are
suspicious, 17.19\% are safe, and
78.89\% are uncertain.
We had, at first, speculated that overflows would be more common among
suspicious instructions, as their bounds, inferred statically, go beyond the
limits of their declared types.
However, our experiments did not let us confirm this hypothesis.
To check the correctness of our approach, we have instrumented the safe instructions,
but have not observed any overflow caused by them.

@@

\begin{figure}[t!]
\begin{center}
\includegraphics[width=\columnwidth]{images/spec_runtime}
\end{center}
\caption{\label{fig:spec_runtime}
Comparison between execution times with and without prunning, 
normalized by the original program's execution time.}
\end{figure}

@@

\noindent
\textbf{Dynamic Detection of Integer Overflows: }
Brumley {\em et al.}~\cite{Brumley07} have developed a tool, RICH, to
secure C programs against integer overflows.
The author's approach consists in instrumenting every integer operation that
might cause an overflow, underflow, or data loss.
The main result of Brumley {\em et al.} is the verification that guarding programs
against integer overflows does not compromise their
performance significantly: the average slowdown across four large applications
is 5\%.
RICH, Brumley {\em et al}'s tool, uses specific features of the x86 architecture
to reduce the instrumentation overhead.
Chinchani {\em et al.}~\cite{Chinchani04} follow a similar approach.
In this work, the authors describe each arithmetic operation formally, and then
use characteristics of the computer architecture to detect overflows at
runtime.
% TODO bibtex: cite Archerr: Runtime environment driven program safety.
Contrary to these previous work, we instrument programs at LLVM's intermediate
representation level, which is machine independent.
Nevertheless, the performance of the programs that we instrument is on par with
Brumley's, even without the support of the static range analysis.

@@

\noindent
\textbf{Static Detection of Integer Overflows: }
Zhang {\em et al.}~\cite{Zhang09} have used static analysis to sanitize
programs against integer overflow based vulnerabilities.
They instrument integer operations in paths from a source to a sink.
In Zhang {\em et al.}'s context, sources are functions that read values from
users, and sinks are memory allocation operations.
Thus, contrary to our work, Zhang {\em et al.}'s only need to instrument
about 10\% of the integer operations in the program.
However, they do not use any form of range analysis to limit the number of
checks inserted in the transformed code.
Wang {\em et a.}~\cite{Wang09} have implemented a tool, IntScope, that combines symbolic execution and taint analysis to detect integer overflow vulnerabilities. The authors have been able to use this tool to successfully identify many vulnerabilities in industrial quality software. Our work, and Wang {\em et a.}'s work are essentially different: they use symbolic execution, whereas we rely on
range analysis.
Contrary to us, they do not transform the program to prevent or detect such event dynamically.
Still in the field of symbolic execution, Molnar
{\em et al.}~\cite{Molnar09} have implemented a tool, SmartFuzz, that analyzes
Linux x86 binaries to find integer overflow bugs.
They prove the existence of bugs by generating test cases for them.

>>>>

\title{A fast and Low-Overhead Technique to Secure Programs Against
Integer Overflows}
%\subtitle{A case study of the Dinamica EGO toolkit}

\authorinfo{Raphael Ernani Rodrigues, Victor Hugo Sperle Campos,
Fernando Magno Quint\~{a}o Pereira}
           {Department of Computer Science -- The Federal University of Minas gerais (UFMG) -- Brazil}
           {\{raphael,victorsc,fernando\}@dcc.ufmg.br}

@@

% Why there is still the need to devise efficient methods to eliminate them.
Programming languages such as Ada or Lisp can be customized to throw exceptions
whenever integer overflows are detected.
Furthermore, there exist recent work proposing to instrument binaries derived
from C, C++ and Java programs to detect the occurrence of overflows
dynamically~\cite{Brumley07,Dietz12}.
Thus, if an overflow happens, then the instrumented program can take some
action, such as to log the event, or to terminate the program.
However, this safety comes with a price: arithmetic operations need to be
surveilled, and the runtime checks cost time.
Zhang {\em et al.}~\cite{Zhang10} have proposed to eliminate some of this
overhead via a tainted flow analysis.
We have a similar goal, yet, our approach is substantially different.

% Our contribution: range analysis.
This paper describes the range analysis algorithm that
we have developed to eliminate overflow checks in instrumented programs.
As we show in Section~\ref{sec:range}, our algorithm has three core insights.
Firstly, in Section~\ref{sub:micro} we propose a three-phase approach to extract
information from comparisons between variables, e.g., $x < y$.
Previous algorithms either deal with these comparisons via
expensive relational analyses~\cite{Cousot78,Lakhdar11,Mine06}, or only
consider comparisons between variables and
constants~\cite{Mahlke01,Patterson95,Stephenson00}.
Secondly, in Section~\ref{sub:prop} we show how we rely on strongly connected
components to achieve speed and precision.
It is well-known that this technique is effective in speeding up constraint
resolution~\cite[Sec 6.3]{Nielson99}.
However, given our three-phase approach, we improve precision
by solving strong components in topological order.
Finally, in Section~\ref{sub:splitting} we propose a new program representation that
is more precise than other intermediate forms used to solve range
analysis sparsely.
This new live range splitting strategy is only valid if the instrumented
program terminates whenever an integer overflow is detected.
If we cannot rely on this guarantee, then our more conservative live range splitting
strategy produces the program representation that Bodik
{\em et al.}~\cite{Bodik00} call Extended Static Single Assignment
form.
In Section~\ref{sub:range} we show that an interprocedural implementation of our
algorithm analyzes programs with half-a-million lines of code in less than ten
seconds.
Furthermore, the speed and memory consumption of our range analysis grows linearly
with the program size.

@@

% An example of vulnerability.
Figure~\ref{fig:ex_buffer_overflow} illustrates an example of integer
overflow vulnerability.
This vulnerability allows an adversary to perform a buffer overrun attack
on a program that is apparently guarded against this type of exploit.
For simplicity, we will use the C datatype \texttt{char}, which represents
8-bit long numbers.
Examples of actual vulnerabilities, with larger primitive types, are given
by Brumley {\em et al.}~\cite{Brumley07}, Dietz {\em et al.}~\cite{Dietz12}, and
Zhang {\em et al.}~\cite{Zhang10}.
The function \texttt{read\_matrix} reads a matrix of bytes from a file.
This matrix is stored as a linear sequence of bytes in a vector \texttt{out}.
The first and second bytes in the file, e.g., \texttt{w} and \texttt{h},
determine the number of rows and columns in the matrix.
If the \texttt{char} datatype could represent arbitrarily long numbers, than this
function would be safe against buffer overruns, because it would only write data
on allocated memory.
However, if, for instance, we have that \texttt{w}$= 16$ and \texttt{h}$= 16$,
then \texttt{s}$= 0$, as pointed before.
In this case, a buffer of length zero would be allocated, and all the data
stored in the file would overwrite memory in the stack of activation records.
By carefully crafting an input file, an adversary can, in this way, overwrite
the return address of \texttt{read\_matrix}, diverting program's execution to
one of \texttt{libc}'s function, such as \texttt{telnet}, for instance.

@@

Notice that $(\infty -\infty)$ is not well-defined.
From the lattice $\varset{Z}$ we define the product lattice
$\varset{Z}^2$ as follows:
%
\begin{equation*}
\varset{Z}^2 = \{ \emptyset \} \cup \{[z_1, z_2] | \ z_1,z_2 \in \varset{Z},
\ z_1 \leq z_2, \  -\infty < z_2 \}
\end{equation*}
%
This interval lattice is partially ordered by the subset relation, which we
denote by ``$\sqsubseteq$".
Range intersection, ``$\sqcap$", is defined by:
\[
[a_1, a_2] \sqcap [b_1, b_2] =
\begin{cases}
[\mbox{max}(a_1, b_1), \mbox{min}(a_2, b_2)], \ \mbox{if} \ a_1 \leq b_1 \leq a_2  \\ \ \ \ \ \mbox{or} \ b_1 \leq a_1 \leq b_2 \\
[a_1, a_2] \sqcap [b_1, b_2] = \emptyset, \ \mbox{otherwise}
\end{cases}
\]
And range union, ``$\sqcup$", is given by:
\[
[a_1, a_2] \sqcup [b_1, b_2] = [\mbox{min}(a_1, b_1), \mbox{max}(a_2, b_2)]
\]

Given an interval $\iota = [l, u]$, we let $\lb{\iota} = l$, and
$\ub{\iota} = u$, where $\lb{\iota}$ is the lower bound and $\ub{\iota}$ is
the upper bound of a variable.
We let \varset{V} be a set of constraint variables, and
$I: \varset{V} \mapsto \varset{Z}^2$ a
mapping from these variables to intervals in $\varset{Z}^2$.
Our objective is to solve a constraint system \varset{C}, formed by constraints
such as those seen in Figure~\ref{fig:eval_function}(left).
We let the $\phi$-functions be as defined by Cytron
{\em et al.}~\cite{Cytron91}: they join different variable names into a single
definition.
Figure~\ref{fig:eval_function}(right) defines a valuation function $e$ on the
interval domain.
Armed with these concepts, we define the range analysis problem as follows:

\begin{definition}
\label{def:rcp}
\textsc{Range Analysis Problem} \\
\textbf{Input:} a set \varset{C} of constraints ranging over a set
\varset{V} of variables. \\
\textbf{Output:} a mapping I such that, for any variable
$V \in \varset{V}$, e(V) = I[V].
\end{definition}

@@

% The macro algorithm
Our range analysis algorithm works in a number of steps.
Firstly, we convert the program to an intermediate representation that gives
us subsidies to perform a sparse analysis.
We have tested our algorithm with two different representations, as we discuss
in Section~\ref{sub:splitting}.
Secondly, we extract constraints from the program representation.
Thirdly, we build a constraint graph, following the strategy pointed by
Su and Wagner~\cite{Su05}.
However, contrary to them, in a next phase we find the strongly connected
components in this graph, collapse them into super-nodes, and sort the
resulting digraph topologically.
Finally, for each strong component, we apply a three-phase approach to determine
the ranges of the variables.

\noindent
\textbf{Building the Constraint Graph. }
Given a set $\varset{C}$ of constraints, which define and/or use constraint
variables from a set $\varset{V}$, we build a constraint graph
$G = (\varset{C} \cup \varset{V}, E)$.
The vertices in \varset{C} are the {\em constraint nodes}, and the
nodes in \varset{V} are the {\em variable nodes}.
If $V \in \varset{V}$ is used in constraint $C \in \varset{C}$, then
we create an edge $\overrightarrow{VC}$.
If constraint $C \in \varset{C}$ defines variable $\varset{V} \in V$, then we
create an edge $\overrightarrow{CV}$.
Figure~\ref{fig:ex_graph} shows the constraint graph that we build for the
program in Figure~\ref{fig:ex_eSSA_cgo}(b).
Our algorithm introduces the notion of {\em future ranges}, which we use to extract
range information from comparisons between variables.
In Section~\ref{sub:splitting} we explain how futures are created.
If $V$ is used by constraint $C$ as the input of a future range, then the edge
from $V$ to $C$ represents what Ferrante {\em et al.} call a {\em control
dependence}~\cite[p.323]{Ferrante87}.
We use dashed lines to represent these edges.
All the other edges denote {\em data dependences}~\cite[p.322]{Ferrante87}.
As we will show later, control dependence edges increase the precision of our
algorithm to solve future bounds.

@@

\subsection{A Three-Phase Approach to Solve Strong Components}
\label{sub:micro}

We find the ranges of the variables in each strongly connected component in
three phases.
First we determine the growth pattern of each variable in the component via
widening.
In the second step, once we know in which directions the values stored in each
variable might grow, we replace futures by actual bounds.
Finally, a narrowing phase starting from conditional tests improves the
precision of our results.

\noindent
\textbf{Widening: } we start solving constraints by determining how each
program variable might grow.
For instance, if a variable is only updated by sums with positive numbers, then
it only grows up.
If, instead, a variable is only updated by sums with negative numbers, then it
grows down.
Some variables can also grow in both directions.
We discover these growth patterns by abstractly interpreting the constraints
that constitute the strongly connected component.
We ensure termination via a widening operator.
Our implementation uses {\em jump-set widening}, which is typically used in
range analysis~\cite[p.228]{Nielson99}.
This operator is a generalization of Cousot and Cousot's original widening
operator~\cite{Cousot77}, which we describe below:
%
\begin{equation*}
I[Y] =
\begin{cases}
\mbox{if} \ I[Y] = [\bot, \bot] \ \mbox{then} \ e(Y) \\
\mbox{elif} \ \lb{e(Y)} < \lb{I[Y]}  \ \mbox{and} \ \ub{e(Y)} > \ub{I[Y]} \ \mbox{then} \ [-\infty, \infty] \\
\mbox{elif} \ \lb{e(Y)} < \lb{I[Y]} \ \mbox{then} \ [-\infty, \ub{I[Y]}] \\
\mbox{elif} \ \ub{e(Y)} > \ub{I[Y]} \ \mbox{then} \ [\lb{I[Y]}, \infty]
\end{cases}
\end{equation*}
%
We let $\lb{[l, u]} = l$ and $\ub{[l, u]} = u$.
We let $\bot$ denote non-itialized intervals, so that $[\bot, \bot] \sqcup
[l, u] = [l, u]$.
This operation only happens at $\phi$-nodes, because we evaluate constraints in
topological order.
The map $I$ and the abstract evaluation function $e$ are determined as in
Figure~\ref{fig:eval_function}.
We have an implementation of $e$ for each operation that the
target programming language provides.
Our current LLVM implementation has 18 different instances of $e$, including
signed and unsigned addition, subtraction, multiplication and division, plus
truncation, the bitwise integer operators and $\phi$-functions.

@@

We have classified the bounds estimated by the static analysis into four
categories.
The first category, called $1$, contains tight bounds:
during program execution, the variable has been assigned an upper,
or lower limit, that equals the limit inferred statically.
The second category, called $n$, contains the bounds that are
within twice the value inferred statically.
For instance, if the range analysis estimates that a variable $v$ is in the
range $[0, 100]$, and during the execution the dynamic profiler finds that
its maximum value is $51$, then $v$ falls into this category.
The third category, $n^2$, contains variables whose actual value is within
a quadratic factor of the estimated value.
In our example, $v$'s upper bound would have to be at most $10$ for it to
be in this category.
Finally, the fourth category contains variables whose estimated value lays
outside a quadratic factor of the actual value.
We call this category {\em imprecise}, and it contains mostly the limits that
our static analysis has marked as either $+\infty$ or $-\infty$.
As we see in Figure~\ref{fig:precision}, 54.11\% of the lower limits that
we have estimated statically are exact.
Similarly, 51.99\% of our upper bounds are also tight.
The figure also shows that, on average, 37.39\% of our lower limits are
imprecise, and 35.40\% of our upper limits are imprecise.
This result is on par with those obtained by more costly analysis, such as
Stephenson {\em et al.}'s~\cite{Stephenson00}.

@@

We have executed the instrumented programs of the integer benchmarks of SPEC
2006 CPU to probe the overhead imposed by our instrumentation.
These programs have been executed with their ``test" input sets.
We have not been able to run the binary that LLVM produces for SPEC's
\texttt{gcc} in our environment, even without any of our passes, due to an
incompatible \texttt{ctype.h} header.
In addition, we have not been able to collect the statistics about
the overflows that occurred in SPEC's \texttt{bzip2}, because the log file
was too large.
We verified more than 3.000.000.000  overflows in this program.
Figure~\ref{fig:unprunedTable} shows the percentage of instructions that
we instrument, without the intervention of the range analysis.
The number of instrumented instructions is relatively low, compared to the
total number of instructions, because we only instrument six different
LLVM bitcodes, in a set of 57 opcodes, not counting intrinsics.
Figure~\ref{fig:unprunedTable} also shows how many instructions have caused
overflows.
On the average, 4.90\% of the instrumented sites have caused integer overflows.

@@

%Prunning analysis
\begin{figure}[t!]
\begin{center}
\begin{small}
\begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}}|l|r|r|r|r|r|}
\hline
Benchmark & \#II & \#E & \%(II, E) & \#U &
\%(II, U) \\ \hline
lbm & 1.142 & 4 & 99,65\% & 4 & 99,65\% \\ \hline
milc & 1.573 & 1.070 & 31,98\% & 1.065 & 32,29\% \\ \hline
namd & 2.965 & 2.900 & 2,19\% & 2.900 & 2,19\% \\ \hline
dealII & 37.234 & 29.870 & 19,78\% & 28.779 & 22,71\% \\ \hline
soplex & 3.268 & 2.927 & 10,43\% & 2.897 & 11,35\% \\ \hline
h264ref & 13.548 & 11.342 & 16,59\% & 11.301 & 16,28\% \\ \hline
astar & 903 & 808 & 10,52\% & 806 & 10,74\% \\ \hline
sjeng & 2.483 & 2.354 & 5,20\% & 2.190 & 11,80\% \\ \hline
mcf & 165 & 164 & 0,61\% & 164 & 0,61\% \\ \hline
omnetpp & 1.533 & 1.313 & 14,35\% & 1.313 & 14,35\% \\ \hline
gcc & 16.330 & 15.282 & 6,42\% & 15.110 & 7,47\% \\ \hline
gobmk & 13.947 & 12.563 & 9,92\% & 12.478 & 10,53\% \\ \hline
libquantum & 943 & 820 & 13,04\% & 817 & 13,36\% \\ \hline
bzip2 & 2.350 & 1.966 & 16,34\% & 1.966 & 16,34\% \\ \hline
hmmer & 3.963 & 3.346 & 15,57\% & 3.304 & 16,63\% \\ \hline
Average & 6.823 & 5.779 & 18,17\% & 5.676 & 19,09\% \\ \hline
\end{tabular*}
\end{small}
\end{center}
\caption{\label{fig:prunningTable}
Instrumentation library with support of static range analysis.
\#II: number of instructions that have been instrumented without range
analysis.
\#E: number of instructions instrumented in the e-SSA form program.
\#U: number of instructions instrumented in the u-SSA form program.}
\end{figure}

Figure~\ref{fig:classificationTable} shows how our range analysis
classifies instructions.
Out of all the 102,790 instructions that we have instrumented in SPEC,
3.92\% are suspicious, 17.19\% are safe, and 78.89\% are uncertain.
This means that we found precise bounds to $3.92 + 17.19 = 21.11\%$ of all
the program variables, and that 78.98\% of them are bound to intervals with
at least one unknown limit.
We had, at first, speculated that overflows would be more common among
suspicious instructions, as their bounds, inferred statically, go beyond the
limits of their declared types.
However, our experiments did not let us confirm this hypothesis.
To check the correctness of our approach, we have instrumented the safe instructions,
but have not observed any overflow caused by them.

@@

\begin{figure}[t!]
\begin{center}
\includegraphics[width=\columnwidth]{images/spec_runtime}
\end{center}
\caption{\label{fig:spec_runtime}
Comparison between execution times with and without pruning, 
normalized by the original program's execution time.}
\end{figure}

@@

\noindent
\textbf{Dynamic Detection of Integer Overflows: }
Brumley {\em et al.}~\cite{Brumley07} have developed a tool, RICH, to
secure C programs against integer overflows.
The author's approach consists in instrumenting every integer operation that
might cause an overflow, underflow, or data loss.
The main result of Brumley {\em et al.} is the verification that guarding programs
against integer overflows does not compromise their
performance significantly: the average slowdown across four large applications
is 5\%.
RICH, Brumley {\em et al}'s tool, uses specific features of the x86 architecture
to reduce the instrumentation overhead.
Chinchani {\em et al.}~\cite{Chinchani04} follow a similar approach.
In this work, the authors describe each arithmetic operation formally, and then
use characteristics of the computer architecture to detect overflows at
runtime.
% TODO bibtex: cite Archerr: Runtime environment driven program safety.
Contrary to these previous work, we instrument programs at LLVM's intermediate
representation level, which is machine independent.
Nevertheless, the performance of the programs that we instrument is on par with
Brumley's, even without the support of the static range analysis.
None of the above works use range analysis to eliminate overflow checks.
This analysis, in our case, could eliminate approximately 45\% of the
tests.

@@

\noindent
\textbf{Static Detection of Integer Overflows: }
Zhang {\em et al.}~\cite{Zhang10} have used static analysis to sanitize
programs against integer overflow based vulnerabilities.
They instrument integer operations in paths from a source to a sink.
In Zhang {\em et al.}'s context, sources are functions that read values from
users, and sinks are memory allocation operations.
Thus, contrary to our work, Zhang {\em et al.}'s only need to instrument
about 10\% of the integer operations in the program.
However, they do not use any form of range analysis to limit the number of
checks inserted in the transformed code.
Wang {\em et a.}~\cite{Wang09} have implemented a tool, IntScope, that combines symbolic execution and taint analysis to detect integer overflow vulnerabilities. The authors have been able to use this tool to successfully identify many vulnerabilities in industrial quality software. Our work, and Wang {\em et a.}'s work are essentially different: they use symbolic execution, whereas we rely on
range analysis.
Contrary to us, they do not transform the program to prevent or detect such event dynamically.
Still in the field of symbolic execution, Molnar
{\em et al.}~\cite{Molnar09} have implemented a tool, SmartFuzz, that analyzes
Linux x86 binaries to find integer overflow bugs.
They prove the existence of bugs by generating test cases for them.
